BATMAN

### 2015-07-12 manuscript/ms.tex
draft of section 2

\documentclass[12pt,preprint]{aastex} 
%\documentclass[apjl]{emulateapj}
\usepackage{amsmath}
\usepackage{color}

\newcommand*{\blue}{\textcolor{blue}}
%\slugcomment{Draft version August 26, 2012}

@@

\section{Algorithm}
The light curve for a planet transiting a star can be described by the function $F(x,y)$, where $F$ is the relative flux of the system when the planet's center is a distance $(x,y)$ away from the stellar center.  $F$ is given by:

\begin{equation}
F = 1 - \int_S{I dS}
\end{equation}
where the stellar intensity $I$ is normalized such that the integrated intensity over the whole stellar surface is unity.

For a radially symmetric limb darkening profile, this surface integral can be reduced to a one-dimensional problem. The intensity is just $I(r)$. Everything in terms of $d = \sqrt{x^2+y^2}$.

There are three cases:\\
case 1: the planet center directly overlaps the stellar center ($d = 0$)\\
case 2: the planet eclipses the star but the centers do not directly overlap ($0<d<1$)\\
case 3: the planet does not eclipse the star ($d>=1$)

First case: planet center overlaps stellar center
\begin{equation}
F(0) = 1 - \int_0^{2\pi}{\int_0^{r_p}{I(r)r dr d\theta}}
\end{equation}

For the second case, where the center of the planet does not overlap the center of the star ($0<d<1$), I make use of the formula for the area of two overlapping circles with radii $r$ and $R$, with centers separated by a distance $d$:

\begin{equation}
A(r, R, d) = r^2\arccos{u} + R^2\arccos{v} - 0.5\sqrt{w}
\end{equation}
where
\begin{eqnarray}
u &=& (d^2+r^2-R^2)/(2dr)\\
v &=& (d^2 + R^2 -r^2)/(2dR) \\
w &=& (-d+r+R)(d+r-R)(d-r+R)(d+r+R)
\end{eqnarray}

Using this expression for the overlapping area, I approximate the integral in FIXME as
\begin{equation}
F(d) = 1 - \sum_{i=0}^{n-1}{\bar{I}(r_i, r_{i+1}) [A(r_{i+1}, r_p, d) - A(r_i, r_p, d)]}
\end{equation}

where $r_0 = d - r_p$ and $r_n = \textrm{MIN}(d+r_p, 1)$.  Works for negative $r$ because $\mu$ is still positive.

The area element is illustrated in Figure FIXME.

To make the sum more efficient, I make the distance $r_i+1 - r_i \propto \arccos{r_i}$  and use Simpson's rule to compute intensity over the area.

\section{Performance}

@@

TODO:
check performance of my alg relative to dA = r*dr*dtheta
xx check whether Simpson's rule is actually faster -- IT ISN'T

>>>>

\documentclass[12pt,preprint]{aastex} 
%\documentclass[apjl]{emulateapj}
\usepackage{amsmath}
\usepackage{color}
%\usepackage{hyperref}

\newcommand*{\blue}{\textcolor{blue}}

@@

\section{Algorithm}
\label{section:algorithm}
To calculate the fraction $\delta$ of stellar flux blocked by a transiting planet, one must integrate the sky-projected intensity of the star ($I$) over the area obscured by the disk of the planet ($S$):
\begin{equation}
\delta = \iint_S{I \mathrm{d}S}
\label{eqn:surfaceint}
\end{equation}
where $I$ is normalized such that the integrated intensity over the stellar disk is unity. This expression is valid for any general stellar surface brightness map; however, it is slow to evaluate numerically because the differential area elements must be small ($<10^{-6}$) in order to achieve better than one part per million (ppm) accuracy.

On the other hand, if the stellar intensity profile is radially symmetric, the two-dimensional calculation in Equation\,\ref{eqn:surfaceint} can be reduced to one dimension and sped up greatly with the following algorithm: 
\begin{equation}
\delta = \sum_{i=1}^{n} I\left(\frac{x_i+x_{i-1}}{2}\right) \left[A(x_{i}, r_p, d) - A(x_{i-1}, r_p, d)\right]
\end{equation}
where $I(x)$ is the 1D stellar intensity profile, $r_p$ is the planetary radius, $d$ is the separation of centers between the star and the planet, and $A(x, r_p, d)$ is the area of intersection between two circles of radii $x$ and $r_p$, separated by a distance $d$.  The sum is carried out over the range $x_0 = d - r_p$ to $x_n = \textrm{MIN}(d + r_p, r_s)$, where $r_s$ is the stellar radius.  The intersecting area is given by:

\begin{equation}
A(x, r_p, d) = 
\begin{cases}
x^2\cos^{-1}{u} + r_p^2\cos^{-1}{v} - 0.5\sqrt{w}, & r_p -d < x < r_p + d\\
x^2, & x \le r_p - d\\
r_p^2 & x \ge r_p +d \\
\end{cases}
\end{equation}
where
\begin{eqnarray}
u &=& (d^2+x^2-r_p^2)/(2dx)\\
v &=& (d^2 + r_p^2 -x^2)/(2dr_p) \\
w &=& (-d+x+r_p)(d+x-r_p)(d-x+r_p)(d+x+r_p)
\end{eqnarray}

The advantage of this integration scheme is that the only error introduced is due to approximating the stellar intensity as constant over the differential area element $\Delta A = A(x_i, r_p, d) - A(x_{i-1}, r_p, d)$.  Computation of $\Delta A$ is expensive, but it needs to be calculated relatively few times ($<<10^{6}$ for sub-ppm accuracy). This makes the integration faster than one with a simpler area element (e.g., $\Delta A = \Delta x \Delta y$), that requires a much smaller step size to achieve the same accuracy.

To further optimize the integration, I use a nonuniform step size.  Typical stellar intensity profiles have larger gradients near the limb of the star than at the center (FIXME ref), so I choose a smaller steps at larger $x$ values: $x_i - x_{i-1} \propto \cos^{-1}{x_{i-1}}$.  An alternative choice would be to make the step-size proportional to the intensity gradient; however, this is more expensive to compute and does not save time??.

See Figure\,\ref{fig:integration} for a schematic illustrating the geometry of the integration.

\begin{figure}
\resizebox{\hsize}{!}{\includegraphics{f1.pdf}}
\caption{Schematic illustration of the integration scheme. The star (large black circle; partially visible) has a radius of 1 and is centered in the plane of the sky at ($x$, $y$) = (0, 0). The planet (smaller black circle) is separated from the center of the stellar disk by a distance $d$ (marked by the solid black line).  The star is partitioned into concentric circles (dotted lines) in order to calculate the integral over the planet disk.  A single integration element $\Delta A$ is shaded in orange. The integration step size illustrated here is larger than for a typical calculation for visual clarity.}
\label{fig:integration}
\end{figure}

\section{The \texttt{batman} package}
I wrote a Python package, \texttt{batman}, that implements the algorithm described in \S\,\ref{section:algorithm} to calculate model transit light curves. \texttt{batman} is being developed on GitHub and full documentation is available at \blue{\url{https://github.com/lkreidberg/batman}}.

In brief, \texttt{batman} supports calculation of light curves for the following stellar intensity profiles:

\begin{align}
  I(\mu) &= I_0                                                                         & &\text{(uniform)}             \\
  I(\mu) &= I_0[1 - u_1(1-\mu)]                                                         & &\text{(linear)}              \\
  I(\mu) &= I_0[1 - u_1(1 - \mu) - u_2(1-\mu)^2]                                        & &\text{(quadratic)}           \\
  I(\mu) &= I_0[1 - u_1(1-\mu^{1/2}) - u_2(1- \mu) - u_3(1-\mu^{3/2}) - u_4(1-\mu^2)]   & &\text{(nonlinear)}
\end{align}

as well as the option to define a custom profile.

The code calculates the separation of centers between the star and the planet based on physical parameters of the transit (e.g., planet-to-star radius ratio, planet orbital period). (FIXME add ref to winn)

-parallelized with OpenMP with a user-specified number of processors (automatically detects whether supported and fails gracefully)
- compatible with Python 2-3.
- listed on PyPI

\subsection{Performance}
\subsection{Comparison with other codes}

@@
